---
title: "IPL performance and salary analysis report"
author: "AKASH DAS , MDS202206"
date: "2022-10-28"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## INTRODUCTION
My project contains 4 parts:  

 1. Team wise performance analysis
 2. Individual player wise performance analysis
 3. Salary wise analysis
 4. Star player analysis

## ABOUT THE DATA SET

I am working with 2 datasets, which are:

#### 1. IPL (2008 -2019) Dataset
#### 2. IPL Player Performance and Salary Dataset

IPL (2008 -2019) Dataset contains 2 csv files:
i.deleveries.csv: it contains all the delivery-wise data of all the matches - match id, inning, batting team, bowling team, over number, batsman name,non-striker batsman name, bowler name, runs in that delivery, wickets fallen in that delivery, and the winner of that match.

ii.matches.csv: it contains all the match-wise data of all the matches - id,season,date,team 1, team 2, toss winner, toss decision,result,win by runs,win by wickets,venue,player of the match,umpires

## WORKING VARIABLES


```{r echo=FALSE , results='asis'}

df=data.frame(Variable_Name = c("batting_team","bowling_team","batsman","bowler","total_runs","batsman_runs","season","toss_decision","player_of_match"),
                 Variable_Type = c("Categorical (Nominal)","Categorical (Nominal)","Categorical (Nominal)","Categorical (Nominal)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)","Categorical (Binary)","Categorical (Nominal)"),
                 Variable_Description=c("batsman's team name","bowler's team name","on strike batsman name","bowler name","runs scored by batting team","runs scored by the batsman","IPL season number","bat/field","best player of the match"))

knitr::kable(df)
```
<br />
2. IPL Player Performance and Salary Dataset contains 1 csv file:
salary.csv: it contains all the IPL season wise data of all the players - id, name,year,final price,role,nationality,team,Ent,age,matches,runs,High Score,Ave,Strike Rate,fifties,hundreds,fours,sixes,catches,stumps,wickets,economy,four wickets,five wickets,Indian,specialist,status.

## WORKING VARIABLES

```{r echo=FALSE , results='asis'}

df=data.frame(Variable_Name = c("Name","Year","Final Price","Team","Runs","StrRate","Fifties","Hundreds","Fours","Sixes","Wkts","FourWkts","FiveWkts"),
                 Variable_Type = c("Categorical (Nominal)","Numeric(Discrete)","Numeric(Continuous)","Categorical (Nominal)","Numeric(Discrete)","Numeric(Continuous)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)","Numeric(Discrete)"),Variable_Description=c("Player name","IPL season year","Player's auction price of that season","Player's IPL team name","Runs scored by player","Strike rate=(total run scored/total balls played )*100","No. of 50's scored by the player","No. of 100's scored by the player","No. of 4's scored by the player","No. of 6's scored by the player","No. of wickets taken by the player","No. of times 4 wickets haul taken by the player","No. of times 5 wickets haul taken by the player"))

knitr::kable(df)
```
## THE OBJECTIVE OF THE PROJECT

1. Team wise performance analysis:

   -  Which team preferred batting or fielding after winning the tosses in IPL ?
   -  Which team scored how many boundaries,over-boundaries,centuries and half-centuries in IPL ?
   -  Which team took how many 4-wicket hauls and 5-wicket hauls in IPL?

2. Individual player wise performance analysis:

   -  Performance analysis of player against other teams.
   -  Innings wise performance analysis of player.
   -  Performance analysis of player against the players he played most number of times.

3. Salary wise analysis:

   -  Which players are receiving huge amount of money but performs not so good ?
   -  Which players are receiving low amount of money but performs well ?

4. Star player analysis:
    
   -  Which players received player of the match award most number of times?
   -  Highest run scorers.
   -  Highest wicket tackers.


## ANALYSIS

### 1. Team wise performance analysis:

### 1.i>  Which team preferred batting or fielding after winning the tosses in IPL ?


```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
library(ggplot2)
library(here)

mat=read.csv(file="C:/Users/USER/Desktop/visu_pro/matches2.csv")
salary=read.csv(file="C:/Users/USER/Desktop/visu_pro/salary.csv")
del=read.csv(file="C:/Users/USER/Desktop/visu_pro/deliveries2.csv")
mat=mat[(mat$season>2007 & mat$season<2018),]

toss_stat=mat%>%
  group_by(toss_winner) %>% 
  summarise(freq=n(),toss_decision=toss_decision)
toss_stat=toss_stat[(toss_stat$freq>25),]

ggplot(toss_stat,aes(x=toss_winner,fill=toss_decision))+geom_bar(position = "dodge")+theme(axis.text.x=element_text(angle=90,hjust=1))+labs(y="Toss_Decision_Frequency",x="Team Names",title = "Post Toss Decision(Bat/Field) BY Different Teams",subtitle = "IPL Season : 2008-2017")


```

##### From the above diagram we can easily see the preferences of each teams after winning the toss in IPL(Season : 2008-2017)

### 1.ii> Which team scored how many boundaries,over-boundaries,centuries and half-centuries in IPL ?

```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
library(treemapify)

team_stat=salary %>%
  group_by(Team) %>% 
  summarise(fours=sum(Fours),sixes=sum(Sixes),four_wkt=sum(FourWkts),five_wkt=sum(FiveWkts),centuries=sum(Hundreds),halfcenturies=sum(Fifties),matches=n())
team_stat=team_stat[(team_stat$matches>50),]


team_stat=salary %>%
  group_by(Team) %>% 
  summarise(fours=sum(Fours),sixes=sum(Sixes),four_wkt=sum(FourWkts),five_wkt=sum(FiveWkts),centuries=sum(Hundreds),halfcenturies=sum(Fifties),matches=n())
team_stat=team_stat[(team_stat$matches>50),]

ggplot(team_stat, aes(x = "", y = fours, fill = Team)) +
  geom_col(color = "black") +
  coord_polar(theta = "y")+geom_text(aes(label = fours),position = position_stack(vjust = 0.5)) +
  guides(fill = guide_legend(title = "Team")) +
  theme_void()+labs(title = "Total NO. Of Boundaries Scored BY Different Teams",subtitle = "IPL Season : 2008-2017")

```

##### From the above pie diagram we can see that which team scores how many boundries in IPL(Season : 2008-2017)


#### Total NO. Of Over-Boundaries Scored BY Different Teams will be represented by similler kind of pie diagram in the project.


```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}

ggplot(team_stat, aes(area = halfcenturies, fill = Team,
                      label = paste(Team,halfcenturies , sep = "\n"))) +
  geom_treemap() +
  geom_treemap_text(colour = "white",
                    place = "centre",
                    size = 15) +
  theme(legend.position = "none")+labs(title = "Total NO. Of Half-Centuries Scored BY Different Teams",subtitle = "IPL Season : 2008-2017")

```

##### From the above treemap diagram we can see that which team scores how many half-centuries in IPL(Season : 2008-2017)


#### Total NO. Of Centuries Scored BY Different Teams will be represented by similler kind of treemap  diagram in the project.


### 1.iii> Which team took how many 4-wicket hauls and 5-wicket hauls in IPL?

```{r}
ggplot(team_stat, aes(x = "", y = four_wkt, fill = Team)) +
  geom_col(color = "black") +
  coord_polar(theta = "y")+geom_text(aes(label = four_wkt),position = position_stack(vjust = 0.5)) +
  guides(fill = guide_legend(title = "Team")) +
  theme_void()+labs(title = "Total NO. Of Four Wicket Hauls Taken BY Different Teams",subtitle = "IPL Season : 2008-2017")

```
 
##### From the above pie diagram we can see that which team takes 4 wicket haul how many times in IPL(Season : 2008-2017) 

#### Total NO. Of Five Wicket Hauls Taken BY Different Teams will be represented by similler kind of pie diagram in the project.


### 2. Individual player wise performance analysis:(In this report we only consider one player's performance as an example of original project)
### we choose MS Dhoni to analys a batsman's performance in IPL

#### 2.i>  Performance analysis of player against other teams.
   
```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
ms=del[(del$batsman=="MS Dhoni"),]
ms3=ms %>% 
  group_by(bowling_team)%>% 
  summarise(tot_run=sum(batsman_runs),balls_played=n())%>%
  arrange(-balls_played)
ms3=ms3[(ms3$balls_played>100),]
ms3$strike_rate=ms3$tot_run/ms3$balls_played*100
ggplot(ms3,aes(reorder(bowling_team,-tot_run),tot_run,fill=bowling_team))+geom_col(position = "dodge")+theme(legend.position = "none")+theme(axis.text.x=element_text(angle=90,hjust=1))+labs(x="Opponent Teams",y="Runs Scored",title = "Runs Scored By MS Dhoni Against Different Teams",subtitle = "IPL Season : 2008-2017")

ggplot(ms3,aes(reorder(bowling_team,-strike_rate),strike_rate,fill=bowling_team))+geom_col(position = "dodge")+theme(legend.position = "none")+theme(axis.text.x=element_text(angle=90,hjust=1))+labs(x="Opponent Teams",y="Strike Rate",title = "MS Dhoni's Strike Rate Against Different Teams",subtitle = "IPL Season : 2008-2017")

```

##### The above bar diagrams shows the performance of MS Dhoni against opponent teams in IPL(Season : 2008-2017)

#### 2.ii>  Innings wise performance analysis of player.
 
```{r, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}

ms_inn=ms%>%
  group_by(inning) %>% 
  summarise(runs=sum(batsman_runs))
ms_inn$innings=c("1st_innings","2nd_innings")

ggplot(ms_inn, aes(x = 3, y = runs, fill = innings)) +
  geom_col(color = "black") +
  geom_text(aes(label = runs),
            position = position_stack(vjust = 0.5)) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "GnBu") +
  xlim(c(0.2, 3 + 0.5)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())+labs(title = "Runs Scored By MS Dhoni In Different Innings",subtitle = "IPL Season : 2008-2017")
```


##### The above donut diagram describes us MS Dhoni's batting performence against other teams.
 
#### 2.iii>  Performance analysis of player against the players he played most number of times.


```{r  echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
ms=del[(del$batsman=="MS Dhoni"),]
ms2=ms %>% 
  group_by(bowler)%>% 
  summarise(tot_run=sum(batsman_runs),balls_played=n())%>%
  arrange(-balls_played)
ms2=ms2[c(1:20),]
ms2$strike_rate=ms2$tot_run/ms2$balls_played*100
ggplot(ms2,aes(reorder(bowler,-tot_run),tot_run,fill=bowler))+geom_col(position = "dodge")+theme(legend.position = "none")+theme(axis.text.x=element_text(angle=90,hjust=1))+labs(x="Bowler Names",y="Runs Scored",title = "Runs Scored By MS Dhoni Against Different Bowlers He Played Most No. Of Times",subtitle = "IPL Season : 2008-2017")

ggplot(ms2,aes(strike_rate,tot_run))+geom_point()+geom_text(aes(label=bowler),hjust=1, vjust=1,cex=3)+geom_hline(yintercept =80)+geom_vline(xintercept =120)+labs(x="Strike Rate",y="Runs Scored",title = "MS Dhoni Vs. Different Bowlers He Played Most No. Of Times",subtitle = "IPL Season : 2008-2017")

```

#### The above scatter plot shows performance of MSD against the bowlers he played most.Observe that the bowlers in the 1st quadrant are favourite bowlers of DHONI, he has scored most number of runs against them with high strike rate and he struggles to play the bowlers in 3rd quadrant.


#### Same strategy can be used to analyse the individual perofromance of any bowler.
 
### 3. Salary wise analysis:

#### 3.i> Which batsman is most profitable for his team  ?

```{r}


sal=salary %>%
  group_by(Name) %>% 
  summarise(run=sum(Runs),money=sum(Final.Price),money_per_run=sum(Runs)/sum(Final.Price)*100000) %>%
  arrange( -money_per_run)


sal=sal[(sal$run > 2000),]

ggplot(sal,aes(money_per_run,reorder(Name, money_per_run),fill=Name))+geom_col()+theme(legend.position = "none")+labs(x="Money Per Run(*100000)",y="Batsman Names",title = "Money Per Run of Different Batsman",subtitle = "(minimum 2000 balls played),IPL:2008-2017")
```

##### From the above diagram we can say that batsman on the top are most profitable batsman for his team, they received less amount of money but score well.

#### 3.ii>   Which bowler is most profitable for his team ?


#### Who are the most profitable bowlers for his team will be represented by similler kind of bar diagram in the project.
 
### 4. Star player analysis:
    
#### 4.i>  Which players received player of the match award most number of times?
```{r}
mat_pom=mat %>% 
  group_by(player_of_match)%>% 
  summarise(freq=n())%>%
  arrange(-freq)
mat_pom1=mat_pom[(mat_pom$freq>9),]
ggplot(mat_pom1,aes(freq,reorder(player_of_match,freq),fill=player_of_match))+geom_col()+theme(legend.position = "none")+labs(x="No. Of Times Player Of The Match Awards Received",y="Player Names",title = "Player Of The Match Awards Received By Different Players ",subtitle = "(minimum 10 awards),IPL Season : 2008-2017")
```
   
#### 4.ii> Highest run scorers.
   
   
   
```{r}
bm=del %>% 
  group_by(batsman)%>% 
  summarise(tot_run=sum(batsman_runs),balls_played=n())
bm1=bm[(bm$balls_played>2000),]
ggplot(bm1,aes(fill=batsman,tot_run,reorder(batsman,tot_run)))+geom_col()+theme(legend.position = "none")+labs(x="Total Of Runs Scored",y="Batsman Names",title = "Total Runs Scored By Different Players ",subtitle = "(minimum 2000 runs),IPL Season : 2008-2017")
```
   

##### The above diagram shows the top run scores in IPL Season : 2008-2017 (minimum 2000 runs scored)

#### 4.iii>  Highest wicket tackers.

#### Same strategy can be used to visualize the Highest wicket bowlers


## CONCLUSION:
   
   In this report I have shown the team performances ,salary wise performance, who are the star players in IPL and individual player's performance analysis of a player as an example.However in the main project I will analyse performance of every player using R-Shinny web app.

 















